---
title: "DATA605_HW12"
author: "Gabriel Santos"
date: '2022-11-07'
output: html_document
---


### The attached who.csv dataset contains real-world data from 2008. 
#### The variables.

* Country: name of the country.

* LifeExp: average life expectancy for the country in years.

* InfantSurvival: proportion of those surviving to one year or more.

* Under5Survival: proportion of those surviving to five years or more.

* TBFree: proportion of the population without TB.

* PropMD: proportion of the population who are MDs.

* PropRN: proportion of the population who are RNs.

* PersExp: mean personal expenditures on healthcare in US dollars at average exchange rate.

* GovtExp: mean government expenditures per capita on healthcare, US dollars at average exchange rate.

* TotExp: sum of personal and government expenditures.


#### The libraries
```{r}
library(ggplot2)
library(tidyverse)
```

#### Data Import

```{r}
who <- read.csv('https://raw.githubusercontent.com/GabrielSantos33/DATA605_HW12/main/who.csv')
summary(who)
```

```{r}
head(who)
```
## Question 1

#### Provide a scatterplot of LifeExp~TotExp, and run simple linear regression. Do not transform the variables. Provide and interpret the F statistics, R^2, standard error,and p-values only. Discuss whether the assumptions of simple linear regression met.

#### Data Exploration
```{r}
who.lm <- lm(LifeExp~TotExp, data=who)
summary(who.lm)
```
```{r}

par(mfrow=c(2,2)) #prints out two rows, two columns of plots
plot(who.lm)
```

```{r}
hist(who.lm$residuals)
```

#### Statistic: 65.26 on 1 and 188 DF, P-value: 7.714e-14, given the small pvalue, which is much below 0.05 indicates the model has some level of validity.

#### Multiple R2: 0.2577, Adjusted R2: 0.2537 - The model only accounts for roughly 25% of the data’s variation.

#### Residual standard error: In our example, the total expenditure required can deviate from the true regression line by approximately 15.3795867, on average.

#### Are the assumptions of simple linear regression met?

#### No, the conditions are not met.
```{r}
#scatterplot
lm <- lm(LifeExp ~ TotExp, data = who)
plot(who$TotExp, who$LifeExp, xlab = 'Total Expenditure', ylab = 'Average Life Expectancy', main='Average Life Expectancy vs Expenditure')
abline(lm)
summary(lm)
```

```{r}
#plots
hist(resid(lm), main = "Histogram of Residuals", xlab = "Residuals")
plot(fitted(lm), resid(lm))
qqnorm(lm$residuals)
qqline(lm$residuals)
```

#### By seeing the R^2 value, we can say that the explained variability is very less @25.77%. The p-value suggests a statistically significant correlation between total expenditures and life expectancy, since p<<0.05. Looking at residuals plots it is clear that there is no constant variability and that residuals are not normally distributed. This model is not a good model.

## Question 2

#### Raise life expectancy to the 4.6 power (i.e., LifeExp^4.6). Raise total expenditures to the 0.06 power (nearly a log transform, TotExp^.06). Plot LifeExp^4.6 as a function of TotExp^.06, and r re-run the simple regression model using the transformed variables. Provide and interpret the F statistics, R^2, standard error, and p-values. Which model is "better?"

```{r}
#model2
LifeExp46 <- who$LifeExp**4.6
TotExp06 <- who$TotExp**0.06
lm1 <- lm(LifeExp46 ~ TotExp06, data = who)
summary(lm1)
plot(who$TotExp**0.06, who$LifeExp**4.6, xlab = 'Total Expenditure raised to 0.06 ', ylab = 'Average Life Expectancy raised to 4.6', main='Average Life Expectancy vs Expenditure')
abline(lm1)
```

```{r}
#plots
hist(resid(lm1), main = "Histogram of Residuals", xlab = "Residuals")
plot(fitted(lm1), resid(lm1))
qqnorm(lm1$residuals)
qqline(lm1$residuals)
```

```{r}
who2 <- who %>% 
  mutate(LifeExp2 = LifeExp^4.6,
         TotExp2 = TotExp^.06)
who.lm2 <- lm(LifeExp2~TotExp2, data=who2)
summary(who.lm2)
```
```{r}
ggplot(who2, aes(x =TotExp2 , y = LifeExp2)) +
  geom_point()+
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, col="red")+
  ggtitle("Plot of Life Expectancy by Total Expenditure") +
  xlab("Total Expenditure ^0.06") + ylab("Life Expectancy ^4.6")
```

```{r}
par(mfrow=c(2,2)) 
plot(who.lm2)
```

#### The 2nd model is better than the 1st model. The R^2 is .729, that means the explained variability is 72.9% compared to 25% for the1st model. Looking at residuals plots, variability is fairly constant with a few outliers and distribution of residuals is nearly normal. So  2nd model is better than the 1st model.

## Questions 3

#### Using the results from 3, forecast life expectancy when TotExp^.06 =1.5. Then forecast life expectancy when TotExp^.06=2.5.

#### The equation: 

#### y =  - 736527910 + 620060216*(x)

```{r}
x = 1.5
y <- (-736527910 + 620060216*x)
y
```

```{r}
LE <- function(x)
{y <- -736527910 + 620060216 *(x) 
y <- y^(1/4.6)
y}
LE(1.5)
LE(2.5)
```
#### Life expectancy when TotExp.06 = 1.5 is approximately 63.3
```{r}
x = 2.5
y <- (-736527910 + 620060216*x)
y
```

```{r}
(y)^(1/4.6)
```
### Life expectancy when TotExp.06=2.5 is approximately 86.5

## Question 4

#### Build the following multiple regression model and interpret the F Statistics, R^2, standard error, and p-values. How good is the model?

#### LifeExp = b0+b1 x PropMd + b2 x TotExp +b3 x PropMD x TotExp

```{r}
#model3
who.lm3 <- lm(LifeExp~TotExp + PropMD + PropMD * TotExp, data=who)
summary(who.lm3)
```
```{r}
par(mfrow=c(2,2)) 
plot(who.lm3)
```

```{r}
par(mfrow=c(1,1))
# residuals histogram
hist(who.lm3$residuals, 
     xlab = "Residuals", ylab = "", 
     main = "Histogram of Residuals Distribution")
```

#### Are the assumptions of simple linear regression met?
* Linearity: The relationship between X and the mean of Y is not linear. Based on the Residuals vs. Fitted plot, the the red line exhibits a quadratic relationship and is not linear.

* Homoscedasticity: The variance of residual is not the same for any value of X. The Scale-Location plot shows the residuals are not spread equally along the ranges of predictor.

* Independence: Observations are not independent of each other. Upon examining the Residuals vs. Fitted plot, we can see a correlation between the variables.

* Multivariate Normality: The nearly normal residual condition doesn’t seem to be met based on the histogram of residuals shown below show the residuals are heavily left skew.

#### How good is the model? Are the assumptions of simple linear regression met?

#### The model is not that great given the criteria mentioned above because the conditions are not met

## Question 5

#### Forecast LifeExp when PropMD=.03 and TotExp = 14. Does this forecast seem realistic? Why or why not?

```{r}
b0 <- 6.277 * 10^1
b1 <- 1.497 * 10^3
b2 <- 7.233 * 10^-5
b3 <- -6.026 * 10^-3
#PropMD
x1 <- .03 
#TotExp
x2 <- 14 
```
```{r}
max(who$LifeExp)
```

```{r}
mean(who$LifeExp)
```

```{r}
LifeExp_new <- ( (6.277*10^1) + (1.497*10^3)*.03 + (7.233*10^(-5))*14 - ((6.026*10^(-3))*0.03*14) ) 
LifeExp_new
```

#### This seems like an outlier as life expectancy of 107 does not look correct. The forecasted life expectancy using the linear model from exercise 4 is 107.7. This forecast is unrealistic with the actual data because the maximum life expectancy from the WHO data is 83 and the mean of all life expectancies is 67.4.